### 2024-11-27 analysis/lod/insilico_mosaic_lod.Rmd
Add folders and files to analysis folder

---
title: "In Silico LOD Detection Evaluation"
date: '`r Sys.Date()`'
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(readxl)
library(here)
```

# Objective
Summarize results of in silico experiment to determine the limit of detection for 
mosaic variant detect using the strelka2 somatic variant caller.

# Methods
- Mixtures of 300 X Illumina reads
    - Chromosome 20 only
    - AFs 
        - 0.00 All HG003
        - 0.01 98:2
        - 0.05 90:10
        - 0.10 80:20
        - 0.25 50:50
        - 0.5 All HG002
- Variant calling using strelka2, performed using the strelka2 app on DNAnexus
- Benchmarking with happy (vcf post processing - TODO add script reference) happy run using pFDA app
- Characterizing results
- Manual Curation
- Nuances and potential issues with using strelka2 for mosaic variant detection.

# Results
```{r}
insilico_lod_df <- read_excel(here("analysis/insilico_lod/Mixture Summary SNPs 01072020.xlsx"),
                              .name_repair = "universal")

tidy_lod_df <- insilico_lod_df %>%
    filter(AF != -1) %>%
    select(-starts_with("SNP")) %>%
    rename(
        fnsnp_pass_count = FN.SNP.PASS,
        fnsnp_pass_pct =  FN.SNP.PASS..,
        fnsnp_all_count = FN.SNP.ALL,
        fnsnp_all_pct = FN.SNP.ALL..,
        recall_pass_val = PASS.Metric.Recall,
        recall_all_val = All.Metric.Recall
    ) %>%
    pivot_longer(
        cols = c(
            fnsnp_pass_count,
            fnsnp_all_count,
            fnsnp_pass_pct,
            fnsnp_all_pct,
            recall_pass_val,
            recall_all_val
        ),
        names_to = c("metric", "var_filter", "met_type"),
        names_sep = "_",
        values_to = "value"
    ) 
```

- AF 0.1 were consistently detected by strelka2 and most variants were detected with AF 0.05. 
- Will start with the assumption that we are able to detect mosaic variants in the RM DNA down to 10% AF using strelka2.
- Can use duplex seq data to validate candidate and putative mosaic variants then optimize the filter parameters is appropriate.
```{r}

ggplot(tidy_lod_df) + 
    geom_vline(aes(xintercept = 0.1), linetype = 1) +
    geom_vline(aes(xintercept = 0.05), linetype = 2) +
    geom_point(aes(x = AF, y = value, fill = var_filter), shape = 21) + 
    facet_wrap(~metric*met_type, ncol = 1, scales = "free_y") + 
    theme_bw()

```

# Session Information
## System Information
```{r}
sessioninfo::platform_info()
```

## Package Versions
```{r}
sessioninfo::package_info() %>% 
    filter(attached = TRUE) %>% 
    select(package, loadedversion, date, source) %>%
    knitr::kable(booktabs = TRUE, row.names = FALSE)
```

### 2025-03-31 analysis/manuscript_figures.Rmd
cleaning up figure Rmarkdown

---
title: "Mosaic_Benchmark_Figures"
author: "Ade, Camille, and Nate"
date: "`r Sys.Date()`"
output: 
    bookdown::html_document2:
        toc: true
        toc_float: true
        df_print: paged
        code_folding: hide
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(here)
library(patchwork)
library(viridis)
library(ggpubr)
library(ggupset)
```

## Load and Tidy Input Data

### Variant VAFs DF for Figure 2
```{r}
## Loading manually curated variants data
chrom_order <- paste0("chr", c(1:22, "X", "Y"))
curated_variants <- read_tsv(here("data", "igv_curated_variants_RM.tsv")) %>%
    arrange(AF) %>%
    mutate(
        index = row_number(),
        CHROM = factor(CHROM, levels = chrom_order)
    ) %>%
    arrange(CHROM, POS, REF, ALT)

curated_variants_longer <- pivot_longer(curated_variants,
    cols = c("BGI_mut_freq", "Element_mut_freq", "Pacbio_mut_freq", "Illumina_mut_freq"),
    names_to = "Method",
    values_to = "VAF"
) %>%
    mutate(Method = str_replace(Method, "_mut_freq", "")) %>%
    mutate(Method = gsub("Illumina", "Illumina 300X", Method)) %>%
    mutate(Method = gsub("BGI", "BGI 100X", Method)) %>%
    mutate(Method = gsub("Element", "Element 136X", Method)) %>%
    mutate(Method = gsub("Pacbio", "Pacbio 108X", Method))

common_variants <- read_tsv(here("data", "RM_VAF_fig_70cand_updated.tsv")) %>%
    arrange(AF) %>%
    mutate(index = row_number())

curated_variants_longer_pacbio_sep <- read_tsv(here("data", "curated_variants_longer_pacbio_sep.tsv"))
```

### In Silico LOD Data
```{r}
insilico_lod_df <- list.files(here("analysis/lod/strelka2_chrom20_mixtures/all_summaries"), full.names = TRUE) %>%
    set_names(., str_extract(., "(?<=af)[:digit:]*(?=_result)")) %>%
    map_dfr(read_csv, .id = "AF") %>%
    mutate(AF = as.numeric(AF))
```

## Figures

### Figure 2
```{r}
annot_df <- data.frame(
    Variant_Category = "**Excluded** from Benchmark",
    l_xmin = -0.25,
    l_xmax = 25,
    l_ymin = -0.03,
    l_ymax = 0.05,
    r_xmin = 113,
    r_xmax = 137,
    r_ymin = -0.03,
    r_ymax = 0.05,
    l_text = "Lack of 300X support",
    l_textx = 6,
    l_texty = 0.4,
    r_text = "Lack of long read support",
    r_textx = 109,
    r_texty = 0.4,
    l_arrowx = -1.5,
    l_arrowy = 0.36,
    l_arrowxend = 1,
    l_arrowyend = 0.02,
    r_arrowx = 110,
    r_arrowy = 0.36,
    r_arrowxend = 116,
    r_arrowyend = 0.03
)

ortho_vafs_df <- curated_variants_longer %>%
    mutate(Variant_Category = case_when(
        RM_KEEP_REMOVE == "KEEP" ~ "Mosaic Benchmark",
        RM_KEEP_REMOVE != "KEEP" ~ "Excluded from Benchmark",
        TRUE ~ "ERROR"
    )) %>%
    mutate(Variant_Category = gsub(
        "((Excluded|Mosaic Benchmark))",
        "**\\1**",
        Variant_Category
    ))

(curated_ortho_vafs_fig <- ggplot(
    curated_variants_longer,
    aes(x = index, y = VAF)
) +
    annotate("rect",
        xmin = -Inf, xmax = Inf,
        ymin = 0.05, ymax = 0.3,
        alpha = 0.4, fill = "gray"
    ) +
    geom_line(
        data = ortho_vafs_df,
        aes(group = index), linewidth = 0.8
    ) +
    geom_point(
        data = ortho_vafs_df,
        aes(fill = Method), size = 5, shape = 21, color = "grey40"
    ) +
    scale_color_viridis() +
    theme_bw() +
    xlab("Ordered by Strelka2 VAF") +
    ylab("Sequencing Method VAF") +
    theme(
        axis.text = element_text(size = 14),
        plot.title = element_text(size = 22),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22),
        axis.text.x = element_blank(),
        legend.title = element_text(size = 19),
        legend.text = element_text(size = 19),
        legend.position = "bottom", legend.box = "vertical"
    ) +
    theme(plot.title = element_text(hjust = 0.5)) +
    facet_wrap(
        ~ factor(Variant_Category,
            levels = c("**Mosaic Benchmark**", "**Excluded** from Benchmark")
        ),
        ncol = 1
    ) +
    theme(
        strip.text = ggtext::element_markdown(size = 20),
        legend.text = ggtext::element_markdown(size = 19)
    ) +
    geom_text(data = annot_df, aes(x = r_textx, y = r_texty, label = r_text), size = 6) +
    geom_text(data = annot_df, aes(x = l_textx, y = l_texty, label = l_text), size = 6) +
    geom_segment(
        data = annot_df,
        aes(x = l_arrowx, y = l_arrowy, xend = l_arrowxend, yend = l_arrowyend),
        arrow = arrow(length = unit(0.3, "cm"), type = "closed")
    ) +
    geom_segment(
        data = annot_df,
        aes(x = r_arrowx, y = r_arrowy, xend = r_arrowxend, yend = r_arrowyend),
        arrow = arrow(length = unit(0.3, "cm"), type = "closed")
    )
)
```

```{r}
ggsave(
    filename = here("figures", "fig2_curated-variants.pdf"),
    plot = curated_ortho_vafs_fig,
    height = 6, width = 11
)
```

### Figure 3

#### Curated Variants VAF

```{r mosaic_VAF_plot, message = FALSE}
(curated_variants_hist <- curated_variants %>%
    mutate(combined_VAF = x_ci_no_Dup / n_ci_no_Dup) %>%
    ggplot(aes(
        x = combined_VAF,
        color = RM_CURATION,
        fill = RM_CURATION
    )) +
    geom_histogram(color = "black", position = "identity", bins = 40) +
    geom_vline(xintercept = c(0.05, 0.3), linetype = "dashed", color = "black") +
    scale_color_manual(values = c("#00BFC4", "#F8766D")) +
    scale_fill_manual(
        values = c("#00BFC4", "#F8766D"),
        labels = c("Mosaic Benchmark (85)", "Excluded from Benchmark (50)")
    ) +
    labs(
        x = "Variant Allele Frequency (VAF)",
        y = "Count"
    ) +
    theme_light() +
    theme(
        legend.title = element_blank(), legend.position = "bottom",
        strip.text = element_blank()
    ) +
    facet_wrap(~RM_CURATION,
        ncol = 2,
        labeller = labeller(RM_CURATION = c(
            "MOSAIC" = "Mosaic Benchmark",
            "NOT MOSAIC" = "Excluded from Benchmark"
        ))
    )
)
```

#### Genomic Context Upset

```{r mosaic_variant_upset_plot, message = FALSE}
# Create upset plot using GRCh38 genomic context annotations from curated variants df
(upset_plot <- curated_variants %>%
    mutate(
        lowmappability = if_else(easy_to_map == "hit", FALSE, TRUE),
        not_lowmappability = if_else(easy_to_map == "hit", TRUE, FALSE),
        homopolymer = if_else(homopolymer_region == "hit", TRUE, FALSE),
        nonhomopolymer = if_else(homopolymer_region == "hit", FALSE, TRUE)
    ) %>%
    group_by(CHROM, POS, REF, ALT, RM_CURATION) %>%
    pivot_longer(
        cols = c(not_lowmappability, lowmappability, homopolymer, nonhomopolymer),
        names_to = "genomic_context",
        values_to = "hit"
    ) %>%
    filter(hit) %>%
    mutate(RM_CURATION = factor(RM_CURATION, levels = c("NOT MOSAIC", "MOSAIC"))) %>%
    summarise(genomic_context = list(genomic_context)) %>%
    ggplot(aes(x = genomic_context, fill = RM_CURATION)) +
    geom_bar(color = "black") +
    scale_x_upset() +
    labs(x = "Genomic Context", y = "Count") +
    scale_fill_discrete(labels = c("Excluded from Benchmark", "Mosaic Benchmark")) +
    theme_bw() +
    theme(legend.title = element_blank(), legend.position = "none"))
```

#### Create panel containing VAF and upset plot

```{r histo_upset panel}
(histo_upset_panel <- curated_variants_hist /
    upset_plot +
    plot_annotation(tag_levels = "a") &
    theme(plot.tag = element_text(face = "bold")) &
    plot_layout(guides = "collect") +
        theme(legend.position = "none"))
```

```{r}
ggsave(
    plot = histo_upset_panel,
    filename = here("figures", "fig3_curated-variants.pdf"),
    height = 6, width = 11
)
```

### Figure 4

```{r}
# Bringing in the Non-RM data and selecting only Element and Pacbio results
non_RM_curated_variants <- curated_variants_longer_pacbio_sep %>%
    select(
        CHROM, POS, REF, ALT, AF,
        Element_depth, Element_alt_depth, Element_map_qual, Element_base_qual,
        Pacbio_revio_depth, Pacbio_revio_alt_depth, Pacbio_revio_map_qual, Pacbio_revio_base_qual,
        Pacbio_sequel_depth, Pacbio_sequel_alt_depth, Pacbio_sequel_map_qual, Pacbio_sequel_base_qual,
        Method, VAF
    ) %>%
    filter(Method %in% c("Element", "Pacbio_revio", "Pacbio_sequel")) %>%
    mutate(Source = "Non RM") %>%
    mutate(Element_mutfreq_nonRM = Element_alt_depth / Element_depth) %>%
    mutate(Pacbio_revio_mutfreq_nonRM = Pacbio_revio_alt_depth / Pacbio_revio_depth) %>%
    mutate(Pacbio_sequel_mutfreq_nonRM = Pacbio_sequel_alt_depth / Pacbio_sequel_depth) %>%
    select(
        CHROM, POS, REF, ALT, AF,
        Element_mutfreq_nonRM,
        Pacbio_revio_mutfreq_nonRM,
        Pacbio_sequel_mutfreq_nonRM
    )

common_variants_pacbio_element_sep <- pivot_longer(common_variants,
    cols = c(
        "BGI_mut_freq", "Element_standard_mut_freq", "Element_long_mut_freq",
        "Pacbio_revio_mut_freq", "Pacbio_sequel_mut_freq", "Illumina_mut_freq"
    ),
    names_to = "Method",
    values_to = "VAF"
) %>%
    mutate(Method = str_replace(Method, "_mut_freq", "")) %>%
    mutate(Method = gsub("Illumina", "Illumina 300X", Method))

rm_vs_nonrm_pairwise <- common_variants_pacbio_element_sep %>%
    select(
        CHROM, POS, REF, ALT, AF,
        Element_depth, Element_alt_depth, Element_map_qual, Element_base_qual,
        Element_standard_depth, Element_standard_alt_depth, Element_standard_map_qual, Element_standard_base_qual,
        Element_long_depth, Element_long_alt_depth, Element_long_map_qual, Element_long_base_qual,
        Pacbio_revio_depth, Pacbio_revio_alt_depth, Pacbio_revio_map_qual, Pacbio_revio_base_qual,
        Pacbio_sequel_depth, Pacbio_sequel_alt_depth, Pacbio_sequel_map_qual, Pacbio_sequel_base_qual,
        Method, VAF
    ) %>%
    filter(Method %in% c("Element", "Element_standard", "Element_long", "Pacbio_revio", "Pacbio_sequel")) %>%
    mutate(Source = "RM") %>%
    mutate(Element_standard_mutfreq_RM = Element_standard_alt_depth / Element_standard_depth) %>%
    mutate(Element_long_mutfreq_RM = Element_long_alt_depth / Element_long_depth) %>%
    mutate(Element_mutfreq_RM = Element_alt_depth / Element_depth) %>%
    mutate(Pacbio_revio_mutfreq_RM = Pacbio_revio_alt_depth / Pacbio_revio_depth) %>%
    mutate(Pacbio_sequel_mutfreq_RM = Pacbio_sequel_alt_depth / Pacbio_sequel_depth) %>%
    select(
        CHROM, POS, REF, ALT, AF,
        Element_mutfreq_RM, Element_standard_mutfreq_RM, Element_long_mutfreq_RM,
        Pacbio_revio_mutfreq_RM, Pacbio_sequel_mutfreq_RM
    ) %>%
    left_join(
        non_RM_curated_variants,
        by = c("CHROM", "POS", "REF", "ALT")
    ) %>%
    select(-AF.x) %>%
    rename(AF = AF.y) %>%
    distinct() %>%
    arrange(AF) %>%
    mutate(index = row_number())

rm_vs_nonrm_pairwise_preprint <- rm_vs_nonrm_pairwise %>%
    select(Element_mutfreq_RM, Element_mutfreq_nonRM) %>%
    rename(mutfreq_RM = Element_mutfreq_RM, mutfreq_nonRM = Element_mutfreq_nonRM) %>%
    mutate(Method = "Element") %>%
    rbind(rm_vs_nonrm_pairwise %>%
        select(Pacbio_revio_mutfreq_RM, Pacbio_revio_mutfreq_nonRM) %>%
        rename(mutfreq_RM = Pacbio_revio_mutfreq_RM, mutfreq_nonRM = Pacbio_revio_mutfreq_nonRM) %>%
        mutate(Method = "Pacbio Revio"))

rm_vs_nonrm_pairwise_preprint_Element <- rm_vs_nonrm_pairwise_preprint %>%
    filter(Method == "Element") %>%
    mutate(mutation = row_number())

rm_vs_nonrm_pairwise_preprint_Pacbio_Revio <- rm_vs_nonrm_pairwise_preprint %>%
    filter(Method == "Pacbio Revio") %>%
    mutate(mutation = row_number())

wilcox_rm_vs_nonrm_pairwise_preprint_Element <- rm_vs_nonrm_pairwise_preprint_Element %>%
    rename("non-RM" = mutfreq_nonRM, "RM" = mutfreq_RM) %>%
    pivot_longer(cols = c("non-RM", "RM"), names_to = "Sample Source", values_to = "VAF")

wilcox_rm_vs_nonrm_pairwise_preprint_Pacbio_Revio <- rm_vs_nonrm_pairwise_preprint_Pacbio_Revio %>%
    rename("non-RM" = mutfreq_nonRM, "RM" = mutfreq_RM) %>%
    pivot_longer(
        cols = c("non-RM", "RM"),
        names_to = "Sample Source",
        values_to = "VAF"
    )

wilcox_rm_vs_nonrm_pairwise_preprint <- rbind(wilcox_rm_vs_nonrm_pairwise_preprint_Element, wilcox_rm_vs_nonrm_pairwise_preprint_Pacbio_Revio)
```

```{r}
(RMvsNonRM_paired_wilcox_preprint <- ggpaired(wilcox_rm_vs_nonrm_pairwise_preprint,
    x = "Sample Source", y = "VAF", id = "mutation",
    color = "Method", line.color = "gray", line.size = 0.4,
    palette = c("#7CAE00", "#C77CFF"),
    facet.by = "Method"
) +
    stat_compare_means(paired = TRUE, size = 6) +
    ylim(0, 0.4) +
    ylab("Variant Allele Frequency (VAF)") +
    xlab("Sample Source") +
    theme(
        plot.title = element_text(hjust = 0.5, size = 22),
        legend.position = "none",
        axis.text = element_text(size = 14),
        axis.title.x = element_text(size = 22),
        axis.title.y = element_text(size = 22),
        strip.text = element_text(size = 16),
        axis.text.x = element_text(size = 16),
        axis.text.y = element_text(size = 16)
    ))
```

```{r}
ggsave(
    plot = RMvsNonRM_paired_wilcox_preprint,
    filename = here("figures", "fig4_VAF-RM-v-nonRM.pdf")
)
```

### Figure S1

```{r}
(lod_fig <- insilico_lod_df %>%
    ggplot() +
    geom_vline(aes(xintercept = 10), linetype = 1) +
    geom_vline(aes(xintercept = 5), linetype = 2) +
    geom_path(aes(x = AF, y = METRIC.Recall, color = Filter, linetype = Type)) +
    geom_point(aes(x = AF, y = METRIC.Recall, fill = Filter, shape = Type)) +
    scale_shape_manual(values = c(21, 22)) +
    theme_bw() +
    labs(x = "Allele Frequency (%)", y = "Recall") +
    theme(legend.position = "bottom"))
```

```{r}
ggsave(
    plot = lod_fig,
    filename = here("figures", "figS1_insilico-lod.pdf")
)
```

### Figure S4
```{r}
(curated_vafs_hist <- ggplot(curated_variants_longer, aes(x = VAF, fill = Method)) +
    geom_histogram(color = "black", alpha = 0.7) +
    facet_wrap(Method ~ ., scales = "fixed", ncol = 1) +
    scale_fill_manual("Method",
        breaks = c("BGI 100X", "Element 136X", "Illumina 300X", "Pacbio 108X"),
        values = c("#F9746D", "#7CAE00", "#00BEC4", "#C77CFF")
    ) +
    theme_bw() +
    theme(legend.position = "none") +
    xlab("Variant Allele Frequency (VAF)") +
    ylab("Number of Manually Curated Variants"))
```

```{r}
ggsave(
    plot = curated_vafs_hist,
    filename = here("figures", "figS4_curated-VAFs.pdf"),
    height = 8, width = 6
)
```

### Figure S5
```{r}
(bench_vafs_hist <- curated_variants_longer %>%
    filter(RM_KEEP_REMOVE == "KEEP") %>%
    ggplot(aes(x = VAF, fill = Method)) +
    geom_histogram(color = "black", alpha = 0.7) +
    facet_wrap(Method ~ ., scales = "fixed", ncol = 1) +
    scale_fill_manual("Method",
        breaks = c("BGI 100X", "Element 136X", "Illumina 300X", "Pacbio 108X"),
        values = c("#F9746D", "#7CAE00", "#00BEC4", "#C77CFF")
    ) +
    theme_bw() +
    theme(legend.position = "none") +
    xlab("Variant Allele Frequency (VAF)") +
    ylab("Number of Mosaic Variants"))
```

```{r}
ggsave(
    plot = bench_vafs_hist,
    filename = here("figures", "figS5_benchmark-VAFs.pdf"),
    height = 8, width = 6
)
```

# Session Information
## System Information

```{r}
sessioninfo::platform_info()
```

## Package Versions
```{r}
sessioninfo::package_info() %>%
    filter(attached = TRUE) %>%
    select(package, loadedversion, date, source) %>%
    knitr::kable(booktabs = TRUE, row.names = FALSE)
```

### 2025-04-03 analysis/manuscript_figures.Rmd
WIP: figure revisions, updated ortho comparison

@@
 chrom_order <- paste0("chr", c(1:22, "X", "Y"))
-curated_variants <- read_tsv(here("data", "igv_curated_variants_RM.tsv")) %>%
-    arrange(AF) %>%
+curated_variants <- read_tsv(
+    here("data", "igv_curated_variants_RM.tsv")
+) %>%
+    dplyr::arrange(AF) %>%
     mutate(
@@
     ) %>%
-    arrange(CHROM, POS, REF, ALT)
+    dplyr::arrange(CHROM, POS, REF, ALT)
+
+benchmark_vars <- curated_variants %>%
+    filter(RM_KEEP_REMOVE == "KEEP") %>%
+    select(CHROM, POS, REF, ALT)
 
 curated_variants_longer <- pivot_longer(curated_variants,
-    cols = c("BGI_mut_freq", "Element_mut_freq", "Pacbio_mut_freq", "Illumina_mut_freq"),
+    cols = c(
+        "BGI_mut_freq",
+        "Element_mut_freq",
+        "Pacbio_mut_freq",
+        "Illumina_mut_freq"
+    ),
     names_to = "Method",
@@
 curated_variants_longer_pacbio_sep <- read_tsv(here("data", "curated_variants_longer_pacbio_sep.tsv"))
+
+benchmark_source_comp_df <- read_tsv(
+    here("data", "ortho_support", "combined_bamread_count.tsv")
+) %>%
+    right_join(benchmark_vars)
 ```
@@
 ```{r}
-insilico_lod_df <- list.files(here("analysis/lod/strelka2_chrom20_mixtures/all_summaries"), full.names = TRUE) %>%
+insilico_lod_df <- list.files(here("data/insilico_lod"), full.names = TRUE) %>%
     set_names(., str_extract(., "(?<=af)[:digit:]*(?=_result)")) %>%
@@
 ```{r}
-annot_df <- data.frame(
-    Variant_Category = "**Excluded** from Benchmark",
-    l_xmin = -0.25,
-    l_xmax = 25,
-    l_ymin = -0.03,
-    l_ymax = 0.05,
-    r_xmin = 113,
-    r_xmax = 137,
-    r_ymin = -0.03,
-    r_ymax = 0.05,
-    l_text = "Lack of 300X support",
-    l_textx = 6,
-    l_texty = 0.4,
-    r_text = "Lack of long read support",
-    r_textx = 109,
-    r_texty = 0.4,
-    l_arrowx = -1.5,
-    l_arrowy = 0.36,
-    l_arrowxend = 1,
-    l_arrowyend = 0.02,
-    r_arrowx = 110,
-    r_arrowy = 0.36,
-    r_arrowxend = 116,
-    r_arrowyend = 0.03
-)
-
 ortho_vafs_df <- curated_variants_longer %>%
     mutate(Variant_Category = case_when(
-        RM_KEEP_REMOVE == "KEEP" ~ "Mosaic Benchmark",
-        RM_KEEP_REMOVE != "KEEP" ~ "Excluded from Benchmark",
+        RM_KEEP_REMOVE == "KEEP" ~ "**Included** in Benchmark",
+        RM_KEEP_REMOVE != "KEEP" ~ "**Excluded** from Benchmark",
         TRUE ~ "ERROR"
-    )) %>%
-    mutate(Variant_Category = gsub(
-        "((Excluded|Mosaic Benchmark))",
-        "**\\1**",
-        Variant_Category
     ))
@@
 (curated_ortho_vafs_fig <- ggplot(
-    curated_variants_longer,
+    ortho_vafs_df,
     aes(x = index, y = VAF)
@@
     ) +
-    geom_line(
-        data = ortho_vafs_df,
-        aes(group = index), linewidth = 0.8
+    geom_line(aes(group = index), linewidth = 0.5) +
+    geom_point(aes(fill = Method), shape = 21, color = "grey40") +
+    scale_color_viridis() +
+    guides(fill = guide_legend(
+        nrow = 1,
+        position = "bottom",
+        direction = "horizontal",
+        title = "Sequencing Method"
+    )) +
+    labs(x = "Strelka2 VAF", y = "VAF") +
+    theme_bw() +
+    theme(
+        strip.text = ggtext::element_markdown(),
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()
     ) +
-    geom_point(
-        data = ortho_vafs_df,
-        aes(fill = Method), size = 5, shape = 21, color = "grey40"
-    ) +
-    scale_color_viridis() +
-    theme_bw() +
-    xlab("Ordered by Strelka2 VAF") +
-    ylab("Sequencing Method VAF") +
-    theme(
-        axis.text = element_text(size = 14),
-        plot.title = element_text(size = 22),
-        axis.title.x = element_text(size = 22),
-        axis.title.y = element_text(size = 22),
-        axis.text.x = element_blank(),
-        legend.title = element_text(size = 19),
-        legend.text = element_text(size = 19),
-        legend.position = "bottom", legend.box = "vertical"
-    ) +
-    theme(plot.title = element_text(hjust = 0.5)) +
     facet_wrap(
-        ~ factor(Variant_Category,
-            levels = c("**Mosaic Benchmark**", "**Excluded** from Benchmark")
-        ),
+        ~Variant_Category,
         ncol = 1
-    ) +
-    theme(
-        strip.text = ggtext::element_markdown(size = 20),
-        legend.text = ggtext::element_markdown(size = 19)
-    ) +
-    geom_text(data = annot_df, aes(x = r_textx, y = r_texty, label = r_text), size = 6) +
-    geom_text(data = annot_df, aes(x = l_textx, y = l_texty, label = l_text), size = 6) +
-    geom_segment(
-        data = annot_df,
-        aes(x = l_arrowx, y = l_arrowy, xend = l_arrowxend, yend = l_arrowyend),
-        arrow = arrow(length = unit(0.3, "cm"), type = "closed")
-    ) +
-    geom_segment(
-        data = annot_df,
-        aes(x = r_arrowx, y = r_arrowy, xend = r_arrowxend, yend = r_arrowyend),
-        arrow = arrow(length = unit(0.3, "cm"), type = "closed")
     )
@@
     plot = curated_ortho_vafs_fig,
-    height = 6, width = 11
+    height = 6, width = 8
 )
@@
 
-#### Curated Variants VAF
-
-```{r mosaic_VAF_plot, message = FALSE}
+```{r currated_VAF, message = FALSE}
 (curated_variants_hist <- curated_variants %>%
@@
     ggplot(aes(
-        x = combined_VAF,
-        color = RM_CURATION,
-        fill = RM_CURATION
+        x = combined_VAF
     )) +
-    geom_histogram(color = "black", position = "identity", bins = 40) +
-    geom_vline(xintercept = c(0.05, 0.3), linetype = "dashed", color = "black") +
-    scale_color_manual(values = c("#00BFC4", "#F8766D")) +
-    scale_fill_manual(
-        values = c("#00BFC4", "#F8766D"),
-        labels = c("Mosaic Benchmark (85)", "Excluded from Benchmark (50)")
+    geom_vline(
+        xintercept = c(0.05, 0.3),
+        linetype = "dashed", color = "black"
+    ) +
+    geom_histogram(
+        color = "black",
+        position = "identity", bins = 40, na.rm = TRUE
     ) +
@@
         x = "Variant Allele Frequency (VAF)",
-        y = "Count"
+        y = "# of Variants"
     ) +
-    theme_light() +
+    theme_bw() +
     theme(
-        legend.title = element_blank(), legend.position = "bottom",
-        strip.text = element_blank()
+        strip.text = ggtext::element_markdown()
     ) +
     facet_wrap(~RM_CURATION,
-        ncol = 2,
+        ncol = 1,
         labeller = labeller(RM_CURATION = c(
-            "MOSAIC" = "Mosaic Benchmark",
-            "NOT MOSAIC" = "Excluded from Benchmark"
+            "MOSAIC" = "**Included** in Benchmark",
+            "NOT MOSAIC" = "**Excluded** from Benchmark"
         ))
@@
 
-#### Genomic Context Upset
-
-```{r mosaic_variant_upset_plot, message = FALSE}
-# Create upset plot using GRCh38 genomic context annotations from curated variants df
-(upset_plot <- curated_variants %>%
-    mutate(
-        lowmappability = if_else(easy_to_map == "hit", FALSE, TRUE),
-        not_lowmappability = if_else(easy_to_map == "hit", TRUE, FALSE),
-        homopolymer = if_else(homopolymer_region == "hit", TRUE, FALSE),
-        nonhomopolymer = if_else(homopolymer_region == "hit", FALSE, TRUE)
-    ) %>%
-    group_by(CHROM, POS, REF, ALT, RM_CURATION) %>%
-    pivot_longer(
-        cols = c(not_lowmappability, lowmappability, homopolymer, nonhomopolymer),
-        names_to = "genomic_context",
-        values_to = "hit"
-    ) %>%
-    filter(hit) %>%
-    mutate(RM_CURATION = factor(RM_CURATION, levels = c("NOT MOSAIC", "MOSAIC"))) %>%
-    summarise(genomic_context = list(genomic_context)) %>%
-    ggplot(aes(x = genomic_context, fill = RM_CURATION)) +
-    geom_bar(color = "black") +
-    scale_x_upset() +
-    labs(x = "Genomic Context", y = "Count") +
-    scale_fill_discrete(labels = c("Excluded from Benchmark", "Mosaic Benchmark")) +
-    theme_bw() +
-    theme(legend.title = element_blank(), legend.position = "none"))
-```
-
-#### Create panel containing VAF and upset plot
-
-```{r histo_upset panel}
-(histo_upset_panel <- curated_variants_hist /
-    upset_plot +
-    plot_annotation(tag_levels = "a") &
-    theme(plot.tag = element_text(face = "bold")) &
-    plot_layout(guides = "collect") +
-        theme(legend.position = "none"))
-```
-
 ```{r}
 ggsave(
-    plot = histo_upset_panel,
-    filename = here("figures", "fig3_curated-variants.pdf"),
-    height = 6, width = 11
+    plot = curated_variants_hist,
+    filename = here("figures", "fig3_curated-variants-VAF.pdf"),
+    height = 6, width = 4
 )
@@
 ```{r}
-# Bringing in the Non-RM data and selecting only Element and Pacbio results
-non_RM_curated_variants <- curated_variants_longer_pacbio_sep %>%
-    select(
-        CHROM, POS, REF, ALT, AF,
-        Element_depth, Element_alt_depth, Element_map_qual, Element_base_qual,
-        Pacbio_revio_depth, Pacbio_revio_alt_depth, Pacbio_revio_map_qual, Pacbio_revio_base_qual,
-        Pacbio_sequel_depth, Pacbio_sequel_alt_depth, Pacbio_sequel_map_qual, Pacbio_sequel_base_qual,
-        Method, VAF
+(source_vaf_comp <- benchmark_source_comp_df %>%
+    filter(
+        str_detect(platform, "element|pacbio"),
+        !str_detect(platform, "long"),
+        !str_detect(platform, "sequel")
     ) %>%
-    filter(Method %in% c("Element", "Pacbio_revio", "Pacbio_sequel")) %>%
-    mutate(Source = "Non RM") %>%
-    mutate(Element_mutfreq_nonRM = Element_alt_depth / Element_depth) %>%
-    mutate(Pacbio_revio_mutfreq_nonRM = Pacbio_revio_alt_depth / Pacbio_revio_depth) %>%
-    mutate(Pacbio_sequel_mutfreq_nonRM = Pacbio_sequel_alt_depth / Pacbio_sequel_depth) %>%
-    select(
-        CHROM, POS, REF, ALT, AF,
-        Element_mutfreq_nonRM,
-        Pacbio_revio_mutfreq_nonRM,
-        Pacbio_sequel_mutfreq_nonRM
+    mutate(tech = str_remove(platform, "_.*")) %>%
+    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
+    summarise(VAF = sum(alt_count) / sum(depth)) %>%
+    mutate(
+        idx = paste0(CHROM, POS, REF, ALT),
+        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
+        tech = factor(tech, levels = c("element", "pacbio"), labels = c("Element", "PacBio Revio"))
+    ) %>%
+    ggpaired(
+        x = "dna_source", fill = "tech",
+        y = "VAF",
+        id = "idx",
+        facet.by = "tech",
+        line.color = "gray",
+        line.size = 0.4,
+        palette = c("#7CAE00", "#C77CFF"),
+    ) +
+    stat_compare_means(paired = TRUE) +
+    ylim(0, 0.4) +
+    theme(legend.position = "none")
 )
-
-common_variants_pacbio_element_sep <- pivot_longer(common_variants,
-    cols = c(
-        "BGI_mut_freq", "Element_standard_mut_freq", "Element_long_mut_freq",
-        "Pacbio_revio_mut_freq", "Pacbio_sequel_mut_freq", "Illumina_mut_freq"
-    ),
-    names_to = "Method",
-    values_to = "VAF"
-) %>%
-    mutate(Method = str_replace(Method, "_mut_freq", "")) %>%
-    mutate(Method = gsub("Illumina", "Illumina 300X", Method))
-
-rm_vs_nonrm_pairwise <- common_variants_pacbio_element_sep %>%
-    select(
-        CHROM, POS, REF, ALT, AF,
-        Element_depth, Element_alt_depth, Element_map_qual, Element_base_qual,
-        Element_standard_depth, Element_standard_alt_depth, Element_standard_map_qual, Element_standard_base_qual,
-        Element_long_depth, Element_long_alt_depth, Element_long_map_qual, Element_long_base_qual,
-        Pacbio_revio_depth, Pacbio_revio_alt_depth, Pacbio_revio_map_qual, Pacbio_revio_base_qual,
-        Pacbio_sequel_depth, Pacbio_sequel_alt_depth, Pacbio_sequel_map_qual, Pacbio_sequel_base_qual,
-        Method, VAF
-    ) %>%
-    filter(Method %in% c("Element", "Element_standard", "Element_long", "Pacbio_revio", "Pacbio_sequel")) %>%
-    mutate(Source = "RM") %>%
-    mutate(Element_standard_mutfreq_RM = Element_standard_alt_depth / Element_standard_depth) %>%
-    mutate(Element_long_mutfreq_RM = Element_long_alt_depth / Element_long_depth) %>%
-    mutate(Element_mutfreq_RM = Element_alt_depth / Element_depth) %>%
-    mutate(Pacbio_revio_mutfreq_RM = Pacbio_revio_alt_depth / Pacbio_revio_depth) %>%
-    mutate(Pacbio_sequel_mutfreq_RM = Pacbio_sequel_alt_depth / Pacbio_sequel_depth) %>%
-    select(
-        CHROM, POS, REF, ALT, AF,
-        Element_mutfreq_RM, Element_standard_mutfreq_RM, Element_long_mutfreq_RM,
-        Pacbio_revio_mutfreq_RM, Pacbio_sequel_mutfreq_RM
-    ) %>%
-    left_join(
-        non_RM_curated_variants,
-        by = c("CHROM", "POS", "REF", "ALT")
-    ) %>%
-    select(-AF.x) %>%
-    rename(AF = AF.y) %>%
-    distinct() %>%
-    arrange(AF) %>%
-    mutate(index = row_number())
-
-rm_vs_nonrm_pairwise_preprint <- rm_vs_nonrm_pairwise %>%
-    select(Element_mutfreq_RM, Element_mutfreq_nonRM) %>%
-    rename(mutfreq_RM = Element_mutfreq_RM, mutfreq_nonRM = Element_mutfreq_nonRM) %>%
-    mutate(Method = "Element") %>%
-    rbind(rm_vs_nonrm_pairwise %>%
-        select(Pacbio_revio_mutfreq_RM, Pacbio_revio_mutfreq_nonRM) %>%
-        rename(mutfreq_RM = Pacbio_revio_mutfreq_RM, mutfreq_nonRM = Pacbio_revio_mutfreq_nonRM) %>%
-        mutate(Method = "Pacbio Revio"))
-
-rm_vs_nonrm_pairwise_preprint_Element <- rm_vs_nonrm_pairwise_preprint %>%
-    filter(Method == "Element") %>%
-    mutate(mutation = row_number())
-
-rm_vs_nonrm_pairwise_preprint_Pacbio_Revio <- rm_vs_nonrm_pairwise_preprint %>%
-    filter(Method == "Pacbio Revio") %>%
-    mutate(mutation = row_number())
-
-wilcox_rm_vs_nonrm_pairwise_preprint_Element <- rm_vs_nonrm_pairwise_preprint_Element %>%
-    rename("non-RM" = mutfreq_nonRM, "RM" = mutfreq_RM) %>%
-    pivot_longer(cols = c("non-RM", "RM"), names_to = "Sample Source", values_to = "VAF")
-
-wilcox_rm_vs_nonrm_pairwise_preprint_Pacbio_Revio <- rm_vs_nonrm_pairwise_preprint_Pacbio_Revio %>%
-    rename("non-RM" = mutfreq_nonRM, "RM" = mutfreq_RM) %>%
-    pivot_longer(
-        cols = c("non-RM", "RM"),
-        names_to = "Sample Source",
-        values_to = "VAF"
-    )
-
-wilcox_rm_vs_nonrm_pairwise_preprint <- rbind(wilcox_rm_vs_nonrm_pairwise_preprint_Element, wilcox_rm_vs_nonrm_pairwise_preprint_Pacbio_Revio)
-```
-
-```{r}
-(RMvsNonRM_paired_wilcox_preprint <- ggpaired(wilcox_rm_vs_nonrm_pairwise_preprint,
-    x = "Sample Source", y = "VAF", id = "mutation",
-    color = "Method", line.color = "gray", line.size = 0.4,
-    palette = c("#7CAE00", "#C77CFF"),
-    facet.by = "Method"
-) +
-    stat_compare_means(paired = TRUE, size = 6) +
-    ylim(0, 0.4) +
-    ylab("Variant Allele Frequency (VAF)") +
-    xlab("Sample Source") +
-    theme(
-        plot.title = element_text(hjust = 0.5, size = 22),
-        legend.position = "none",
-        axis.text = element_text(size = 14),
-        axis.title.x = element_text(size = 22),
-        axis.title.y = element_text(size = 22),
-        strip.text = element_text(size = 16),
-        axis.text.x = element_text(size = 16),
-        axis.text.y = element_text(size = 16)
-    ))
 ```
@@
 ggsave(
-    plot = RMvsNonRM_paired_wilcox_preprint,
+    plot = source_vaf_comp,
     filename = here("figures", "fig4_VAF-RM-v-nonRM.pdf")

### 2025-04-03 scripts/readcount_parser/src/main.rs

@@ -1,0 +1,169 @@
+use clap::Parser;
+use csv::ReaderBuilder;
+use std::collections::HashMap;
+use std::error::Error;
+use std::fs::File;
+use std::io::{BufRead, BufReader, Write};
+use std::path::Path;
+
+/// Command-line arguments structure
+#[derive(Parser, Debug)]
+#[command(version, about, long_about = None)]
+struct Args {
+    /// Path to the file containing the list of bam-readcount output files
+    #[arg(short, long)]
+    bam_readcount_list: String,
+
+    /// Path to the variant list file
+    #[arg(short, long)]
+    variants: String,
+
+    /// Path to the output file
+    #[arg(short, long)]
+    output: String,
+}
+
+/// Reads the variant list from a file and stores it in a HashMap
+fn read_variants(
+    variant_file: &str,
+) -> Result<HashMap<(String, u32), (String, String)>, Box<dyn Error>> {
+    let file = File::open(variant_file)?;
+    let mut rdr = ReaderBuilder::new()
+        .has_headers(true) // Indicates that the file has headers
+        .delimiter(b'\t')  // Specifies tab-delimited fields
+        .from_reader(file);
+
+    let mut variants = HashMap::new();
+
+    for result in rdr.records() {
+        let record = result?;
+        let chr = record.get(0).ok_or("Missing chromosome field")?.to_string();
+        let pos: u32 = record
+            .get(1)
+            .ok_or("Missing position field")?
+            .parse()
+            .map_err(|_| "Invalid position value")?;
+        let ref_base = record.get(2).ok_or("Missing ref_base field")?.to_string();
+        let alt_base = record.get(3).ok_or("Missing alt_base field")?.to_string();
+        variants.insert((chr, pos), (ref_base, alt_base));
+    }
+    Ok(variants)
+}
+
+/// Extracts dna_source and platform from the file path
+fn extract_metadata(file_path: &str) -> Option<(String, String)> {
+    let path = Path::new(file_path);
+    let components: Vec<_> = path.components().collect();
+    if components.len() >= 3 {
+        let dna_source = components[components.len() - 2]
+            .as_os_str()
+            .to_string_lossy()
+            .to_string();
+        let filename = path.file_name()?.to_string_lossy();
+        let platform = filename.split("_HG002").next()?.to_string();
+        Some((dna_source, platform))
+    } else {
+        None
+    }
+}
+
[…]

### 2025-04-04 analysis/manuscript_figures.Rmd
revised ortho analysis including statistical tests

@@
 library(ggupset)
+library(glmmTMB)
+library(broom)
 ```
@@
 
+Preparing data and statistical tests
+
 ```{r}
-(source_vaf_comp <- benchmark_source_comp_df %>%
+## Preparing data
+vaf_dat <- benchmark_source_comp_df %>%
     filter(
         str_detect(platform, "element|pacbio"),
-        !str_detect(platform, "long"),
-        !str_detect(platform, "sequel")
+        !str_detect(platform, "long|sequel")
+    ) %>%
+    mutate(
+        platform = str_remove(platform, "_.*"),
+        dna_source = factor(dna_source, levels = c("nonRM", "RM")),
+        variant_id = paste0(CHROM, ":", POS, ":", REF, ":", ALT)
+    ) %>%
+    group_by(dna_source, platform, variant_id) %>%
+    summarise(
+        vaf = sum(alt_count) / sum(depth),
+        alt_count = sum(alt_count),
+        ref_count = sum(ref_count),
+        depth = sum(depth)
+    )
+
+# ---- Global GLMM Binomial Model ----
+glmm_global <- glmmTMB(
+    cbind(alt_count, ref_count) ~ dna_source + platform + (1 | variant_id),
+    family = binomial,
+    data = vaf_dat
+)
+
+# Overdispersion check
+od_result <- performance::check_overdispersion(glmm_global)
+
+# Reduced model (no dna_source)
+glmm_null <- update(glmm_global, . ~ . - dna_source)
+lrt_result <- anova(glmm_null, glmm_global)
+
+# Extract LRT p-value
+lrt_pval <- lrt_result$`Pr(>Chisq)`[2]
+
+# ---- Per-Variant Binomial GLMs ----
+variant_tests <- vaf_dat %>%
+    group_by(variant_id) %>%
+    filter(n_distinct(dna_source) == 2) %>%
+    nest() %>%
+    mutate(
+        model = map(data, ~ glm(cbind(alt_count, ref_count) ~ dna_source + platform,
+            family = binomial, data = .x
+        )),
+        tidy = map(model, ~ tidy(., conf.int = TRUE))
+    ) %>%
+    unnest(tidy) %>%
+    filter(term == "dna_sourceRM") %>%
+    mutate(p_adj = p.adjust(p.value, method = "BH"))
+
+# Determine direction
+variant_tests <- variant_tests %>%
+    mutate(direction = case_when(
+        p_adj < 0.05 & estimate > 0 ~ "Higher in RM",
+        p_adj < 0.05 & estimate < 0 ~ "Higher in nonRM",
+        TRUE ~ "Not significant"
+    ))
+
+# Count and get significant variants
+total_sig <- sum(variant_tests$p_adj < 0.05)
+higher_rm <- sum(variant_tests$direction == "Higher in RM")
+higher_nonrm <- sum(variant_tests$direction == "Higher in nonRM")
+
+sig_variants <- variant_tests %>%
+    filter(p_adj < 0.05) %>%
+    arrange(estimate) %>%
+    mutate(row_id = row_number())
+```
+
+Out of `r nrow(variant_tests)` variants tested, `r total_sig` exhibited statistically significant VAF differences (FDR < 0.05) between RM and nonRM samples.
+Of these, `r higher_rm` variants had higher VAFs in RM, while {higher_nonrm} had higher VAFs in nonRM.
+The global GLMM model detected a significant overall difference in VAF profiles between RM and nonRM (LRT p = `r signif(lrt_pval, 3)`).
+Overdispersion was not detected (dispersion ratio = `r signif(od_result$dispersion_ratio, 3)`, p = `r signif(od_result$p_value, 3)`).
+
+#### Forest Plot - Supplemental?
+```{r}
+(forest_plot <- ggplot(
+    sig_variants,
+    aes(x = estimate, y = fct_reorder(variant_id, estimate))
+) +
+    geom_errorbarh(
+        aes(
+            xmin = conf.low,
+            xmax = conf.high,
+            color = direction
+        ),
+        height = 0.3
+    ) +
+    geom_point(color = "black") +
+    geom_vline(xintercept = 0, linetype = "dashed") +
+    labs(
+        x = "Log-odds (RM vs nonRM)", y = "Variant ID",
+        title = "Significantly Different VAFs (FDR < 0.05)"
+    ) +
+    scale_color_manual(values = c(
+        "Higher in RM" = "red",
+        "Higher in nonRM" = "blue"
+    )) +
+    theme_minimal()
+)
+```
+
+Paried Plot
+```{r}
+# Merge direction info into your VAF comparison data
+source_vaf_comp <- benchmark_source_comp_df %>%
+    filter(
+        str_detect(platform, "element|pacbio"),
+        !str_detect(platform, "long|sequel")
     ) %>%
@@
     group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
-    summarise(VAF = sum(alt_count) / sum(depth)) %>%
+    summarise(VAF = sum(alt_count) / sum(depth), .groups = "drop") %>%
     mutate(
-        idx = paste0(CHROM, POS, REF, ALT),
+        idx = paste0(CHROM, ":", POS, ":", REF, ":", ALT),
         dna_source = factor(dna_source, levels = c("nonRM", "RM")),
-        tech = factor(tech, levels = c("element", "pacbio"), labels = c("Element", "PacBio Revio"))
+        tech = factor(tech,
+            levels = c("element", "pacbio"),
+            labels = c("Element", "PacBio Revio")
+        )
     ) %>%
-    ggpaired(
-        x = "dna_source", fill = "tech",
-        y = "VAF",
-        id = "idx",
+    left_join(
+        variant_tests %>%
+            select(variant_id, direction),
+        by = c("idx" = "variant_id")
+    ) %>%
+    mutate(
+        line_color = case_when(
+            direction == "Higher in RM" ~ "red",
+            direction == "Higher in nonRM" ~ "blue",
+            TRUE ~ "lightgrey"
+        )
+    )
+
+# Generate ggpaired plot with custom line coloring
+(ggpaired_plot <- ggpaired(
+    source_vaf_comp,
+    x = "dna_source", y = "VAF", id = "idx",
     facet.by = "tech",
-        line.color = "gray",
-        line.size = 0.4,
-        palette = c("#7CAE00", "#C77CFF"),
+    line.color = "direction", # use mapped line color
+    line.size = 0.4
 ) +
-    stat_compare_means(paired = TRUE) +
     ylim(0, 0.4) +
-    theme(legend.position = "none")
+    theme_minimal() +
+    theme(legend.position = "none") +
+    labs(
+        title = "VAF Comparison by DNA Source and Platform",
+        subtitle = "Line color indicates VAF direction for significant variants (FDR < 0.05)"
     )
-```
-
-```{r}
-ggsave(
-    plot = source_vaf_comp,
-    filename = here("figures", "fig4_VAF-RM-v-nonRM.pdf")
 )

### 2025-04-08 analysis/manuscript_figures.Rmd
updated figures and formatting Rmarkdown

@@
 ---
-title: "Mosaic_Benchmark_Figures"
+title: "mosaic benchmark manuscript figures"
 author: "Ade, Camille, and Nate"
@@
 
-## Load and Tidy Input Data
+# Load and Tidy Input Data
 
-### Variant VAFs DF for Figure 2
+## Variant VAFs DF for Figure 2
 ```{r}
@@
 curated_variants <- read_tsv(
-    here("data", "igv_curated_variants_RM.tsv")
+    here("data", "igv_curated_variants_RM.tsv"),
+    show_col_types = FALSE
 ) %>%
@@
 
-common_variants <- read_tsv(here("data", "RM_VAF_fig_70cand_updated.tsv")) %>%
+common_variants <- read_tsv(
+    here("data", "RM_VAF_fig_70cand_updated.tsv"),
+    show_col_types = FALSE
+) %>%
     arrange(AF) %>%
@@
 
-curated_variants_longer_pacbio_sep <- read_tsv(here("data", "curated_variants_longer_pacbio_sep.tsv"))
+curated_variants_longer_pacbio_sep <- read_tsv(
+    here("data", "curated_variants_longer_pacbio_sep.tsv"),
+    show_col_types = FALSE
+)
 
 benchmark_source_comp_df <- read_tsv(
-    here("data", "ortho_support", "combined_bamread_count.tsv")
+    here("data", "ortho_support", "combined_bamread_count.tsv"),
+    show_col_types = FALSE
 ) %>%
@@
 
-### In Silico LOD Data
+## In Silico LOD Data
 ```{r}
@@
     set_names(., str_extract(., "(?<=af)[:digit:]*(?=_result)")) %>%
-    map_dfr(read_csv, .id = "AF") %>%
+    map_dfr(read_csv, show_col_types = FALSE, .id = "AF") %>%
     mutate(AF = as.numeric(AF))
@@
 
-## Figures
+## RM v. nonRM VAF Comparison
 
-### Figure 2
-```{r}
-ortho_vafs_df <- curated_variants_longer %>%
-    mutate(Variant_Category = case_when(
-        RM_KEEP_REMOVE == "KEEP" ~ "**Included** in Benchmark",
-        RM_KEEP_REMOVE != "KEEP" ~ "**Excluded** from Benchmark",
-        TRUE ~ "ERROR"
-    ))
-
-(curated_ortho_vafs_fig <- ggplot(
-    ortho_vafs_df,
-    aes(x = index, y = VAF)
-) +
-    annotate("rect",
-        xmin = -Inf, xmax = Inf,
-        ymin = 0.05, ymax = 0.3,
-        alpha = 0.4, fill = "gray"
-    ) +
-    geom_line(aes(group = index), linewidth = 0.5) +
-    geom_point(aes(fill = Method), shape = 21, color = "grey40") +
-    scale_color_viridis() +
-    guides(fill = guide_legend(
-        nrow = 1,
-        position = "bottom",
-        direction = "horizontal",
-        title = "Sequencing Method"
-    )) +
-    labs(x = "Strelka2 VAF", y = "VAF") +
-    theme_bw() +
-    theme(
-        strip.text = ggtext::element_markdown(),
-        axis.text.x = element_blank(),
-        axis.ticks.x = element_blank()
-    ) +
-    facet_wrap(
-        ~Variant_Category,
-        ncol = 1
-    )
-)
-```
-
-```{r}
-ggsave(
-    filename = here("figures", "fig2_curated-variants.pdf"),
-    plot = curated_ortho_vafs_fig,
-    height = 6, width = 8
-)
-```
-
-### Figure 3
-
-```{r currated_VAF, message = FALSE}
-(curated_variants_hist <- curated_variants %>%
-    mutate(combined_VAF = x_ci_no_Dup / n_ci_no_Dup) %>%
-    ggplot(aes(
-        x = combined_VAF
-    )) +
-    geom_vline(
-        xintercept = c(0.05, 0.3),
-        linetype = "dashed", color = "black"
-    ) +
-    geom_histogram(
-        color = "black",
-        position = "identity", bins = 40, na.rm = TRUE
-    ) +
-    labs(
-        x = "Variant Allele Frequency (VAF)",
-        y = "# of Variants"
-    ) +
-    theme_bw() +
-    theme(
-        strip.text = ggtext::element_markdown()
-    ) +
-    facet_wrap(~RM_CURATION,
-        ncol = 1,
-        labeller = labeller(RM_CURATION = c(
-            "MOSAIC" = "**Included** in Benchmark",
-            "NOT MOSAIC" = "**Excluded** from Benchmark"
-        ))
-    )
-)
-```
-
-```{r}
-ggsave(
-    plot = curated_variants_hist,
-    filename = here("figures", "fig3_curated-variants-VAF.pdf"),
-    height = 6, width = 4
-)
-```
-
-### Figure 4
-
-Preparing data and statistical tests
+## Loading Data
 
@@
     )
+```
 
+### Statistical Analysis
+
+#### Methods
+
+##### Global Comparison of VAF Profiles
+
+We tested whether the variant allele frequency (VAF) profiles differed globally between DNA sources (RM vs nonRM) using a generalized linear mixed-effects model (GLMM).
+We modeled the number of alternate reads ($\text{alt_count}$) out of total reads ($\text{alt_count}$ + $\text{ref_count}$) per variant per sample using a binomial distribution.
+
+The model included DNA source ($\text{dna_source}$) and sequencing platform ($\text{platform}$) as fixed effects, and variant ID ($\text{variant_id}$) as a random intercept to account for baseline allele frequency heterogeneity.
+The model was fit using the glmmTMB package in R.
+A likelihood ratio test (LRT) was used to compare the full model with a reduced model excluding the dna_source term to test whether VAF profiles differed between RM and nonRM sources.
+
+##### Per-Variant Tests for Directional VAF Differences
+
+To identify individual variants with significant VAF differences between RM and nonRM samples, we fit a separate binomial generalized linear model (GLM) for each variant:
+
+$$
+\text{alt_count} \sim \text{Binomial}(n, p), \quad 
+\log\left(\frac{p}{1 - p}\right) = \beta_0 + \beta_1 \cdot \text{dna_source} + \beta_2 \cdot \text{platform}
+$$
+
+From each model, we extracted the log-odds estimate for dna_source = RM, its 95% confidence interval, and the associated p-value.
+P-values were adjusted across all variants using the Benjamini–Hochberg procedure to control the false discovery rate (FDR).
+
+```{r}
 # ---- Global GLMM Binomial Model ----
@@
 
+# Figures
+
+## Figure 2
+```{r}
+ortho_vafs_df <- curated_variants_longer %>%
+    mutate(Variant_Category = case_when(
+        RM_KEEP_REMOVE == "KEEP" ~ "**Included** in Benchmark",
+        RM_KEEP_REMOVE != "KEEP" ~ "**Excluded** from Benchmark",
+        TRUE ~ "ERROR"
+    ))
+
+(curated_ortho_vafs_fig <- ggplot(
+    ortho_vafs_df,
+    aes(x = index, y = VAF)
+) +
+    annotate("rect",
+        xmin = -Inf, xmax = Inf,
+        ymin = 0.05, ymax = 0.3,
+        alpha = 0.4, fill = "gray"
+    ) +
+    geom_line(aes(group = index), linewidth = 0.5) +
+    geom_point(aes(fill = Method), shape = 21, color = "grey40") +
+    scale_color_viridis() +
+    guides(fill = guide_legend(
+        nrow = 1,
+        position = "bottom",
+        direction = "horizontal",
+        title = "Sequencing Method"
+    )) +
+    labs(x = "Strelka2 VAF", y = "VAF") +
+    theme_bw() +
+    theme(
+        strip.text = ggtext::element_markdown(),
+        axis.text.x = element_blank(),
+        axis.ticks.x = element_blank()
+    ) +
+    facet_wrap(
+        ~Variant_Category,
+        ncol = 1
+    )
+)
+```
+
+```{r}
+ggsave(
+    filename = here("figures", "fig2_curated-variants.pdf"),
+    plot = curated_ortho_vafs_fig,
+    height = 6, width = 8
+)
+
+ggsave(
+    filename = here("figures", "fig2_curated-variants.png"),
+    plot = curated_ortho_vafs_fig,
+    height = 6, width = 8
+)
+```
+
+## Figure 3
+
+```{r currated_VAF, message = FALSE}
+(curated_variants_hist <- curated_variants %>%
+    mutate(combined_VAF = x_ci_no_Dup / n_ci_no_Dup) %>%
+    ggplot(aes(
+        x = combined_VAF
+    )) +
+    geom_vline(
+        xintercept = c(0.05, 0.3),
+        linetype = "dashed", color = "black"
+    ) +
+    geom_histogram(
+        color = "black",
+        position = "identity", bins = 40, na.rm = TRUE
+    ) +
+    labs(
+        x = "Variant Allele Frequency (VAF)",
+        y = "# of Variants"
+    ) +
+    theme_bw() +
+    theme(
+        strip.text = ggtext::element_markdown()
+    ) +
+    facet_wrap(~RM_CURATION,
+        ncol = 1,
+        labeller = labeller(RM_CURATION = c(
+            "MOSAIC" = "**Included** in Benchmark",
+            "NOT MOSAIC" = "**Excluded** from Benchmark"
+        ))
+    )
+)
+```
+
+```{r}
+ggsave(
+    plot = curated_variants_hist,
+    filename = here("figures", "fig3_curated-variants-VAF.pdf"),
+    height = 6, width = 4
+)
+
+ggsave(
+    plot = curated_variants_hist,
+    filename = here("figures", "fig3_curated-variants-VAF.png"),
+    height = 6, width = 4
+)
+```
+
+## Figure 4
+
 Out of `r nrow(variant_tests)` variants tested, `r total_sig` exhibited statistically significant VAF differences (FDR < 0.05) between RM and nonRM samples.
-Of these, `r higher_rm` variants had higher VAFs in RM, while {higher_nonrm} had higher VAFs in nonRM.
 The global GLMM model detected a significant overall difference in VAF profiles between RM and nonRM (LRT p = `r signif(lrt_pval, 3)`).
@@
 
-#### Forest Plot - Supplemental?
-```{r}
-(forest_plot <- ggplot(
-    sig_variants,
-    aes(x = estimate, y = fct_reorder(variant_id, estimate))
-) +
-    geom_errorbarh(
-        aes(
-            xmin = conf.low,
-            xmax = conf.high,
-            color = direction
-        ),
-        height = 0.3
-    ) +
-    geom_point(color = "black") +
-    geom_vline(xintercept = 0, linetype = "dashed") +
-    labs(
-        x = "Log-odds (RM vs nonRM)", y = "Variant ID",
-        title = "Significantly Different VAFs (FDR < 0.05)"
-    ) +
-    scale_color_manual(values = c(
-        "Higher in RM" = "red",
-        "Higher in nonRM" = "blue"
-    )) +
-    theme_minimal()
-)
-```
-
-Paried Plot
+### Paired Plot
 ```{r}
@@
 # Generate ggpaired plot with custom line coloring
-(ggpaired_plot <- ggpaired(
+(source_vaf_plot <- ggpaired(
     source_vaf_comp,
@@
 
-### Figure S1
+```{r}
+ggsave(
+    plot = source_vaf_plot,
+    filename = here("figures", "fig4_sample-source-vaf.pdf")
+)
+
+ggsave(
+    plot = source_vaf_plot,
+    filename = here("figures", "fig4_sample-source-vaf.png")
+)
+```
+
+### Forest Plot
+
+Extra plot do not plan to include in manuscript.
+
+```{r}
+(forest_plot <- ggplot(
+    sig_variants,
+    aes(x = estimate, y = fct_reorder(variant_id, estimate))
+) +
+    geom_errorbarh(
+        aes(
+            xmin = conf.low,
+            xmax = conf.high,
+            color = direction
+        ),
+        height = 0.3
+    ) +
+    geom_point(color = "black", alpha = 0.25) +
+    geom_vline(xintercept = 0, linetype = "dashed") +
+    labs(
+        x = "Log-odds (RM vs nonRM)", y = "Variant ID",
+        title = "Significantly Different VAFs (FDR < 0.05)"
+    ) +
+    scale_color_manual(values = c(
+        "Higher in RM" = "lightgrey",
+        "Higher in nonRM" = "blue"
+    )) +
+    theme_minimal()
+)
+```
+
+## Figure S1
 
@@
 )
+
+ggsave(
+    plot = lod_fig,
+    filename = here("figures", "figS1_insilico-lod.png")
+)
 ```
 
-### Figure S4
+## Figure S4
 ```{r}
@@
 )
+
+ggsave(
+    plot = curated_vafs_hist,
+    filename = here("figures", "figS4_curated-VAFs.png"),
+    height = 8, width = 6
+)
 ```
 
-### Figure S5
+## Figure S5
 ```{r}
@@
 )
+
+ggsave(
+    plot = bench_vafs_hist,
+    filename = here("figures", "figS5_benchmark-VAFs.png"),
+    height = 8, width = 6
+)
 ```

### 2025-04-10 analysis/manuscript_figures.Rmd
figure clean-up and resizing

@@
 (curated_variants_hist <- curated_variants %>%
-    mutate(combined_VAF = x_ci_no_Dup / n_ci_no_Dup) %>%
+    mutate(
+        combined_VAF = x_ci_no_Dup / n_ci_no_Dup,
+        Variant_Category = case_when(
+            RM_KEEP_REMOVE == "KEEP" ~ "**Included** in Benchmark",
+            RM_KEEP_REMOVE != "KEEP" ~ "**Excluded** from Benchmark",
+            TRUE ~ "ERROR"
+        )
+    ) %>%
     ggplot(aes(
@@
     ) +
-    facet_wrap(~RM_CURATION,
-        ncol = 1,
-        labeller = labeller(RM_CURATION = c(
-            "MOSAIC" = "**Included** in Benchmark",
-            "NOT MOSAIC" = "**Excluded** from Benchmark"
-        ))
-    )
+    facet_wrap(~Variant_Category, ncol = 1)
 )
@@
     filename = here("figures", "fig3_curated-variants-VAF.pdf"),
-    height = 6, width = 4
+    height = 4, width = 4
 )
@@
     filename = here("figures", "fig3_curated-variants-VAF.png"),
-    height = 6, width = 4
+    height = 4, width = 4
 )
@@
 
+Median benchmark variant coverage for Element short insert and PacBio Revio data.
+
+```{r}
+benchmark_source_comp_df %>%
+    filter(
+        str_detect(platform, "element|pacbio"),
+        !str_detect(platform, "long|sequel")
+    ) %>%
+    mutate(tech = str_remove(platform, "_.*")) %>%
+    group_by(dna_source, tech, CHROM, POS, REF, ALT) %>%
+    reframe(depth = sum(depth)) %>%
+    group_by(dna_source, tech) %>%
+    reframe(coverage = median(depth))
+```
+
 ### Paired Plot
-```{r}
+```{r message = FALSE, warning = FALSE}
+## custom ggpaired for configurable line.alpha and shape aes
+source(here("scripts", "custom_ggpaired.R"))
+```
+
+```{r message = FALSE}
 # Merge direction info into your VAF comparison data
@@
         variant_tests %>%
-            select(variant_id, direction),
+            select(variant_id, direction, p_adj),
         by = c("idx" = "variant_id")
@@
         line_color = case_when(
+            p_adj > 0.05 ~ "grey20",
             direction == "Higher in RM" ~ "red",
-            direction == "Higher in nonRM" ~ "blue",
-            TRUE ~ "lightgrey"
+            direction == "Higher in nonRM" ~ "darkblue",
+            TRUE ~ "green"
+        ),
+        line_alpha = case_when(
+            p_adj > 0.05 ~ 0.75,
+            direction == "Higher in RM" ~ 1,
+            direction == "Higher in nonRM" ~ 1,
+            TRUE ~ 1
+        ),
+        dna_source = if_else(dna_source == "nonRM",
+            "non-RM", dna_source
         )
@@
     facet.by = "tech",
-    line.color = "direction", # use mapped line color
-    line.size = 0.4
+    point.shape = 21,
+    point.fill = "line_color",
+    point.alpha = "line_alpha",
+    line.color = "line_color",
+    line.alpha = "line_alpha" # use mapped line color
 ) +
     ylim(0, 0.4) +
-    theme_minimal() +
+    scale_color_identity() +
+    scale_fill_identity() +
+    theme_bw() +
     theme(legend.position = "none") +
     labs(
-        title = "VAF Comparison by DNA Source and Platform",
-        subtitle = "Line color indicates VAF direction for significant variants (FDR < 0.05)"
+        x = "Sample Source",
+        y = "VAF"
     )
@@
 ggsave(
-    plot = source_vaf_plot,
+    plot = source_vaf_plot, height = 3, width = 6,
     filename = here("figures", "fig4_sample-source-vaf.pdf")
@@
 ggsave(
-    plot = source_vaf_plot,
+    plot = source_vaf_plot, height = 3, width = 6,
     filename = here("figures", "fig4_sample-source-vaf.png")

### 2025-04-10 scripts/custom_ggpaired.R

@@ -1,0 +1,157 @@
+#'@include utilities.R ggpar.R
+NULL
+#'Plot Paired Data
+#'@description Plot paired data.
+#'@inheritParams ggboxplot
+#'@param cond1 variable name corresponding to the first condition.
+#'@param cond2 variable name corresponding to the second condition.
+#'@param x,y x and y variables, where x is a grouping variable and y contains
+#'  values for each group. Considered only when \code{cond1} and \code{cond2}
+#'  are missing.
+#'@param id variable name corresponding to paired samples' id. Used to connect
+#'  paired points with lines.
+#'@param color points and box plot colors. To color by conditions, use color =
+#'  "condition".
+#'@param fill box plot fill color. To change fill color by conditions, use fill
+#'  = "condition".
+#'@param line.color line color.
+#'@param linetype line type.
+#'@param point.size,line.size point and line size, respectively.
+#'@param point.alpha,line.alpha point and line size, respectively.
+#'@param width box plot width.
+#'@param ... other arguments to be passed to be passed to \link{ggpar}().
+#' @examples
+#'
+#'# Example 1
+#'#::::::::::::::::::::::::::::::::::::::::::
+#' before <-c(200.1, 190.9, 192.7, 213, 241.4, 196.9, 172.2, 185.5, 205.2, 193.7)
+#' after <-c(392.9, 393.2, 345.1, 393, 434, 427.9, 422, 383.9, 392.3, 352.2)
+#'
+#' d <- data.frame(before = before, after = after)
+#' ggpaired(d, cond1 = "before", cond2 = "after",
+#'     fill = "condition", palette = "jco")
+#'
+#'# Example 2
+#'#::::::::::::::::::::::::::::::::::::::::::
+#'ggpaired(ToothGrowth, x = "supp", y = "len",
+#'  color = "supp", line.color = "gray", line.size = 0.4,
+#'  palette = "npg")
+#'
+#'@export
+ggpaired <- function(data, cond1, cond2, x = NULL, y = NULL, id = NULL,
+                     color = "black", fill = "white", palette = NULL,
+                     width = 0.5, 
+                     point.size = 1.2, line.size = 0.5,
+                     point.alpha = 1, line.alpha = 1,
+                     line.color = "black",
+                     line.type = "solid" ,point.shape = 21, point.fill = "white",
+                     title = NULL, xlab = "Condition", ylab = "Value",
+                     facet.by = NULL, panel.labs = NULL, short.panel.labs = TRUE,
[…]

### 2025-05-09 analysis/variant_filtering/mosaic_decision_tree_NIST_RM.Rmd
Add decision tree Rmd

---
title: "<span style='font-size: 18pt'>Decision_Tree_Variant_Filtering</span>"
author: "Camille Daniels"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 5
    toc_float:
      collapsed: true
editor_options: 
  chunk_output_type: inline
---

#### PURPOSE

Filter potential mosaic variant database to get list of variants for manual curation (green squares) and generate final mosaic benchmark vcf and bed files.

This process consists of 4 phases:

1. Compare strelka2 HG002 potential mosaic variants to identify denovo variants

2. Filtering potential mosaic database using decision tree heuristics

3. Get variants for manual curation

4. Benchmark bed exclusion

#### DECISION TREE

![mosaic decision tree](https://github.com/usnistgov/giab-HG002-mosaic-benchmark/blob/main/figures/Suppl_Figure_3_mosaic_decision_tree.pdf)

#### RESOURCES

Mosaic benchmark generation requires the following resources:

- strelka2 vcfeval true positives

- GIAB v4.2.1 HG002/HG003/HG004 trio benchmark harmonized denovo VCF

- HG002 GRCh38 potential mosaic database - ![tsv](https://ftp-trace.ncbi.nlm.nih.gov/ReferenceSamples/giab/release/AshkenazimTrio/HG002_NA24385_son/mosaic_v1.10/GRCh38/SNV/SupplementaryFiles/)

- GIABv4.2.1 HG002 GRCh38 AllTandemRepeatsandHomopolymers_slop5.bed

- human.b38.genome

- bedtools

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(here)
library(data.table)
library(tidyverse)
library(vcfR)
library(patchwork)
#library(bedtoolsr)
library(networkD3)
library(htmlwidgets)
library(htmltools)
library(webshot2)
library(grid)
library(gridExtra)
library(flextable)
library(gt)
library(gtExtras)
library(huxtable)
library(png)
library(scales)
library(chromote)
library(ggupset)
library(UpSetR)
library(ComplexUpset)
```

```{r here, include=FALSE}
#<-- Clean up Rmd using here package - to get rid of absolute paths
here::i_am("mosaic_decision_tree_NIST_RM.Rmd")
```

#### I. Comparison of strelka2 vcfeval TPs vs trio denovo benchmark and annotation

```{r get_mosaic_giabv4.2.1_tp_vcf, results = 'hide'}

# Get vcfeval TP VCF

tp_vcf <- read.vcfR(here("~/git_repos/mosaic-benchmarking/data/strelka2_wgs_parents-vs-child/squash_ploidy_normals_excluded_vcfeval_comparison/tp.vcf.gz"))
tp_vcf_df <- data.frame(CHROM = tp_vcf@fix[, "CHROM"], POS = tp_vcf@fix[, "POS"])

```

```{r mosaic tp vcf count}
print(paste0("HG002 mosaic TRUE positives - variants that match giab v4.2.1 benchmark: ", nrow(tp_vcf)))
```

##### Get trio denovo vcf

```{r get_trio_denovo_vcf, results = 'hide'}

#Read trio denovo VCF

denovo_vcf <- read.vcfR(here("HG002_3_4_GRCh38_1_22_v4.2.1_draft_benchmark_trio_harmonize_w_merged_vcf_output_trioinconsistent_denovo.vcf"))
denovo_vcf_df <- data.frame(CHROM = denovo_vcf@fix[, "CHROM"], POS = denovo_vcf@fix[, "POS"])

```

```{r trio denovo count}

print(paste0("GIABv4.2.1 trio denovo benchmark variants: ", nrow(denovo_vcf)))

```

##### Compare vcfeval tp.vcf to GIAB v4.2.1 trio denovo benchmark

```{r get_mosaic_giabv4.2.1_tp_vcf_and_trio_denovo_vcf}

if (!dir.exists("denovo_compare")) {
  dir.create("denovo_compare")
}

# Merge the draft mosaic benchmark TP VCF and denovo VCF data based on CHROM and POS

denovo_mosaic_tp_hg2_hits<- merge(tp_vcf_df, denovo_vcf_df, by = c("CHROM", "POS"))

fwrite(denovo_mosaic_tp_hg2_hits, file = "denovo_compare/denovo_hits_mosaic_giab4.2.1_tp.tsv", sep = "\t")
write_delim(denovo_mosaic_tp_hg2_hits[, c("CHROM", "POS")], file = "denovo_compare/denovo_list_hits_mosaic_giab4.2.1_tp.tsv", delim = "\t", col_names = TRUE)

#Use list to order output files below
chrom_order <- c("chr1", "chr2", "chr3", "chr4", "chr5", "chr6", "chr7", "chr8", "chr9", "chr10", "chr11", "chr12", 
                "chr13", "chr14", "chr15","chr16","chr17", "chr18", "chr19", "chr20", "chr21", "chr22", "chr23", "chr24")

#Get denovo hits with full denovo database annotation
denovo_mosaic_tp_hits_list <- fread("denovo_compare/denovo_list_hits_mosaic_giab4.2.1_tp.tsv")
denovo_db_df <- fread("~/Documents/mosaic_variants_project/draft_benchmark/denovo_db/denovo_variant_database_indel_fixed_genomic_context.tsv") #df needed for downstream analysis

merge_denovo_db_mosaic_tp_hits <- inner_join(denovo_mosaic_tp_hits_list, denovo_db_df, by = c("CHROM", "POS"))
merge_denovo_db_mosaic_tp_hits_sorted <- merge_denovo_db_mosaic_tp_hits [order(factor(merge_denovo_db_mosaic_tp_hits$CHROM, levels = chrom_order)),]

fwrite(merge_denovo_db_mosaic_tp_hits_sorted, "denovo_compare/denovo_db_hits_mosaic_giabv4.2.1_tp.tsv", sep = "\t")

#Get non-matches from denovo comparison
denovo_mosaic_tp_hg2_no_hits <- anti_join(tp_vcf_df, denovo_vcf_df, by = c("CHROM", "POS"))
fwrite(denovo_mosaic_tp_hg2_no_hits, file = "denovo_compare/denovo_no_hits_mosaic_giabv4.2.1_tp.tsv", sep = "\t")
write_delim(denovo_mosaic_tp_hg2_no_hits[, c("CHROM", "POS")], file = "denovo_compare/denovo_list_no_hits_mosaic_giab4.2.1_tp.tsv", delim = "\t", col_names = TRUE)

# Print counts for matching and non-matching variants
print(paste0("Variants that hit hg2 mosaic tp vcf AND trio denovo v4: ", nrow(denovo_mosaic_tp_hg2_hits)))
print(paste0("Variants that hit hg2 mosaic tp vcf, BUT NOT trio denovo v4: ", nrow(denovo_mosaic_tp_hg2_no_hits)))

```

#### II. FILTERING POTENTIAL MOSAIC DATABASE

```{r get_potential_mosaic_db}

# Get HG002 potential mosaic database - strelka2 vcfeval FPs annotated with orthogonal bamreadcounts and GIAB v4.2.1 genomic stratifications

draft_mosaic_db <- fread("~/Documents/mosaic_variants_project/draft_benchmark/v0.0.5_draft_mosaic_benchmark_db_NIST_RM/mosaic_variant_database_RM.tsv")
print(paste0("v0.0.5 draft mosaic benchmark variants : ", nrow(draft_mosaic_db)))

```

##### Partition potential mosaic variant database into upperCI\<=0.03 and NON upperCI\<=0.03

```{r upperCI_3}

upper_CI3_df <- draft_mosaic_db %>%
  filter(Ortho_NO_DUP_Upper_CI <= 0.03)
  fwrite(upper_CI3_df, "1_upperCI_3_RM.tsv", sep = "\t")

not_upper_CI3_df<- draft_mosaic_db %>%    
  filter(!(Ortho_NO_DUP_Upper_CI <= 0.03) | is.na(Ortho_NO_DUP_Upper_CI)) 
  fwrite(not_upper_CI3_df, "1_not_upperCI_3_RM.tsv", sep = "\t")

print(paste0("NO_DUP_upperCI <=0.03 : ", nrow(upper_CI3_df)))
print(paste0("NOT NO_DUP_upperCI <=0.03 : ", nrow(not_upper_CI3_df)))  
```

##### Partition upperCI\<=0.03 non-matches using n_ci_pacbio filter

```{r n_ci_pacbio}

#Using 0.5% quantile value for mosaic_readcounts_RM$n_ci_no_Dup and mosaic_readcounts_RM$Pacbio_depth filters

n_ci_pacbio_yes_df <- not_upper_CI3_df %>%
  filter(not_upper_CI3_df$n_ci_no_Dup < 138 | not_upper_CI3_df$Pacbio_depth < 52)

n_ci_pacbio_no_df <- not_upper_CI3_df %>%
  filter(!(not_upper_CI3_df$n_ci_no_Dup < 138 | not_upper_CI3_df$Pacbio_depth < 52))

fwrite(n_ci_pacbio_yes_df, "2_n_ci_pacbio_yes_RM.tsv", sep = "\t")
fwrite(n_ci_pacbio_no_df, "2_n_ci_pacbio_no_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03 and YES n_ci_pacbio filter : ", nrow(n_ci_pacbio_yes_df)))
print(paste0("not UpperCI <=0.03 and NO n_ci_pacbio filter : ", nrow(n_ci_pacbio_no_df))) 
```

##### Partition upperCI\<=0.03 non-matches that are NO n_ci_pacbio using germline deletion OVERLAP and NO OVERLAP filter

```{r filter_germline_indel_overlaps}

#Partition remaining variants that overlap germline indels or next to insertions

germline_indel_overlap_df <- n_ci_pacbio_no_df %>%
  filter(Germline_Indel_Check == "Deletion_Overlap" | germline_insertion == "hit")

germline_indel_no_overlap_df <- n_ci_pacbio_no_df %>%
  filter(!(Germline_Indel_Check == "Deletion_Overlap" | germline_insertion == "hit"))

fwrite(germline_indel_overlap_df, "3_germline_indel_overlaps_RM.tsv", sep = "\t")
fwrite(germline_indel_no_overlap_df, "3_germline_indel_NON_overlaps_RM.tsv", sep ="\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, germline indel overlap : ", nrow(germline_indel_overlap_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap : ", nrow(germline_indel_no_overlap_df))) 
```

##### Partition NON germline overlaps using x_ci filter

```{r x_ci}

x_ci_yes_df <- germline_indel_no_overlap_df[x_ci_no_Dup/n_ci_no_Dup > 0.5]
x_ci_no_df <- germline_indel_no_overlap_df[!(x_ci_no_Dup/n_ci_no_Dup) > 0.5]

fwrite(x_ci_yes_df, "4_x_ci_yes_RM.tsv", sep = "\t")              ###exclude from vcf
fwrite(x_ci_no_df,"4_x_ci_no_RM.tsv", sep = "\t")

print(paste0("NOT UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, YES x_ci_filter :", nrow(x_ci_yes_df)))
print(paste0("NOT UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter :", nrow(x_ci_no_df))) 
```

##### Partition germline deletion NO OVERLAP, NO x_ci using easytomap and not_easytomap filter

```{r easy_or_not_easytomap}
####<---- AMEND to add correct df for subset

easytomap_df <- subset(x_ci_no_df, grepl("^Easy",Genomic_Context))
not_easytomap_df <-subset(x_ci_no_df, !grepl("^Easy",Genomic_Context))
fwrite(easytomap_df, "5_easytomap_RM.tsv", sep = "\t" )
fwrite(not_easytomap_df, "5_not_easytomap_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap :", nrow(easytomap_df))) 
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, NOT_easytomap :", nrow(not_easytomap_df)))
```

##### Partition easytomap into lowerCI \>=0.05 and NOT \>=0.05

```{r easytomap_lowerCI5_notlowerCI5}

easy_lowerCI5_df <- easytomap_df %>%
  filter(Ortho_NO_DUP_Lower_CI >= 0.05) 
fwrite(easy_lowerCI5_df, "6_easytomap_lowerCI5_RM.tsv", sep = "\t")       ### variants for manual curation (decision tree green box)

write_delim(easy_lowerCI5_df[, c("CHROM", "POS")], file = "6_list_hits_easytomap_lowerCI5.tsv", delim = "\t", col_names = TRUE)

easy_NOT_lowerCI5_df <- easytomap_df %>%
  filter(!Ortho_NO_DUP_Lower_CI >= 0.05) 
fwrite(easy_NOT_lowerCI5_df, "6_easytomap_not_lowerCI5_RM.tsv", sep = "\t")

write_delim(easy_NOT_lowerCI5_df[, c("CHROM", "POS")], file = "6_list_no_hits_easytomap_lowerCI5_RM.tsv", delim = "\t", col_names = TRUE)

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap >=0.05: ", nrow(easy_lowerCI5_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap NOT >=0.05: ", nrow(easy_NOT_lowerCI5_df))) 
```

##### Partition easytomap, NOT lowerCI \>=0.05 into candidate and putative

```{r easytomap_NOT_lowerCI5_cand_put}

easy_NOT_lowerCI5_cand_df<- easy_NOT_lowerCI5_df %>%
  filter(var_cat == 'candidate')
fwrite(easy_NOT_lowerCI5_cand_df, "7_easytomap_not_lowerCI5_candidate_RM.tsv", sep = "\t")

easy_NOT_lowerCI5_put_df <- easy_NOT_lowerCI5_df %>%
  filter(var_cat != 'candidate')
fwrite(easy_NOT_lowerCI5_put_df, "7_easytomap_not_lowerCI5_putative_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap NOT >=0.05, candidate: ", nrow(easy_NOT_lowerCI5_cand_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap NOT >=0.05, putative: ", nrow(easy_NOT_lowerCI5_put_df)))

```

##### Partition easytomap, NOT lowerCI \>=0.05, putative into pacbio alt depth \>=2 YES or NO filter

```{r easytomap_lowerCI5_cand_put_pacbio_depth}

easy_NOT_lowerCI5_put_pacbio_depth_yes_df <- easy_NOT_lowerCI5_put_df %>%
  filter(Pacbio_alt_depth >= 2)
 fwrite(easy_lowerCI5_df, "8_easytomap_not_lowerCI5_putative_pbio_depth_yes_RM.tsv", sep = "\t")

 easy_NOT_lowerCI5_put_pacbio_depth_no_df <- easy_NOT_lowerCI5_put_df %>%
  filter(!(Pacbio_alt_depth >= 2))
 fwrite(easy_lowerCI5_df, "8_easytomap_not_lowerCI5_putative_pbio_depth_no_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap NOT lowerCI >=0.05, putative, pacbio_alt_depth YES: ", nrow(easy_NOT_lowerCI5_put_pacbio_depth_yes_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, easytomap NOT lowerCI >=0.05, putative, pacbio_alt_depth NO: ", nrow(easy_NOT_lowerCI5_put_pacbio_depth_no_df)))

```

##### Partition NOT_easytomap using homopolymer or NON-homopolymer filter

```{r noteasytomap_hompolymer_non_homopolymer}

not_easy_homopol_df <- not_easytomap_df %>%
  filter(homopolymer_region == "hit")
fwrite(not_easy_homopol_df, "9_not_easy_homopolymer_RM.tsv", sep = "\t")

not_easy_NON_homopol_df <- not_easytomap_df %>%
  filter(homopolymer_region == "homopolymer_region_no_hit")
fwrite(not_easy_NON_homopol_df, "9_not_easy_NON_homopolymer_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, homopolymer: ", nrow(not_easy_homopol_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, NON-homopolymer: ", nrow(not_easy_NON_homopol_df)))

```

##### Partition NOT_easytomap, NON-homopolymer using pacbio lowerCI \>= 0.05 YES or NO filter

```{r noteasytomap_nonhompolymer_pacbio_lowerCI5_not}

#Partition non-homopolymer into Pacbio LowerCI > 0.05 and NOT

not_easy_NON_homopol_pacbio_lowerCI5_df <- not_easy_NON_homopol_df %>%
  filter(Pacbio_Lower_CI >= 0.05) 
  fwrite(not_easy_NON_homopol_pacbio_lowerCI5_df,"10_not_easy_NON_homopolymer_pbio_lowerCI5_RM.tsv", sep = "\t") ### variants for manual curation (decision tree green box)

  write_delim(not_easy_NON_homopol_pacbio_lowerCI5_df[, c("CHROM", "POS")], file = "10_list_hits_not_easy_NON_homopolymer_pbio_lowerCI5_RM.tsv", delim = "\t", col_names = TRUE)

not_easy_NON_homopol_pacbio_NOT_lowerCI5_df <- not_easy_NON_homopol_df %>%
  filter(!Pacbio_Lower_CI >= 0.05)
  fwrite(not_easy_NON_homopol_pacbio_NOT_lowerCI5_df, "10_not_easy_NON_homopolymer_not_pbio_lowerCI5_RM.tsv", sep = "\t")

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, NON-homopolymer, pbio lowerCI >=0.05: ", nrow(not_easy_NON_homopol_pacbio_lowerCI5_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, NON-homopolymer, NOT pbio lowerCI >=0.05: ", nrow(not_easy_NON_homopol_pacbio_NOT_lowerCI5_df)))

```

##### Partition NOT_easytomap, NON-homopolymer, pacbio lowerCI \>= 0.05 FAIL, into pacbio upperCI \< 0.03 YES or NO filter

```{r noteasytomap_nonhompolymer_NOT_pacbio_lowerCI5_pacbio_upperCI3}

not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_upperCI3_df <- not_easy_NON_homopol_pacbio_NOT_lowerCI5_df %>%
   filter(Pacbio_Upper_CI < 0.03) 
   fwrite(not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_upperCI3_df, "11_not_easy_NON_homopolymer_pbio_upperCI3_RM.tsv", sep = "\t") ### exclude vcf only

not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_NOT_upperCI3_df <- not_easy_NON_homopol_pacbio_NOT_lowerCI5_df %>%
   filter(!Pacbio_Upper_CI < 0.03)
   fwrite(not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_NOT_upperCI3_df, "11_not_easy_NON_homopolymer_pbio_not_upperCI3_RM.tsv", sep = "\t") ### exclude vcf and bed

print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, NON-homopolymer, pbio upperCI < 0.03: ", nrow(not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_upperCI3_df)))
print(paste0("not UpperCI <=0.03, NO n_ci_pacbio filter, NO germline indel overlap, NO x_ci_filter, not_easytomap, NON-homopolymer, NOT pbio upperCI < 0.03: ", nrow(not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_NOT_upperCI3_df)))

```

#### III. GET VARIANTS FOR MANUAL CURATION

```{r get_variants_for_manual_curation, message = FALSE}

# Variants for curation correspond to green squares in decision tree

#Join easytomap lower >=5 and noteasytomap non-homopolymer pacbio lower CI > 5 dataframes

variants_to_curate_df <- full_join(easy_lowerCI5_df, not_easy_NON_homopol_pacbio_lowerCI5_df) %>%
  arrange(factor(CHROM, levels = chrom_order), POS, REF, ALT)

fwrite(variants_to_curate_df, "variants_for_igv_curation_RM.tsv", sep = "\t")

```

```{r variants_to_curate}

print(paste0("Variants for IGV curation: ", nrow(variants_to_curate_df)))
```

##### Import manually curated variants and partition into include and exclude for new benchmark vcf

```{r partition_manually_curated_variants, message = FALSE}

#Import tsv of manually curated variants (135)

curated_variants_all_df <- fread("igv_curated_variants_RM.tsv", sep = "\t") %>%
  arrange(factor(CHROM, levels = chrom_order), POS, REF, ALT)

#Mosaic variants to include in NIST RM benchmark vcf - only variants that are MOSAIC in RM_CURATION column

curated_variants_include_vcf_df <- curated_variants_all_df %>%
   filter(RM_CURATION == "MOSAIC")

fwrite(curated_variants_include_vcf_df, "igv_curated_variants_RM_include_vcf.tsv", sep = "\t")
write_delim(curated_variants_include_vcf_df[, c("CHROM", "POS", "REF", "ALT")], file = "igv_curated_variants_RM_include_vcf_list.tsv", delim = "\t", col_names = TRUE)

#Variants to exclude from benchmark vcf
curated_variants_exclude_vcf_df <- anti_join(curated_variants_all_df, curated_variants_include_vcf_df)

fwrite(curated_variants_exclude_vcf_df, "igv_curated_variants_RM_exclude_vcf.tsv", sep = "\t")
write_delim(curated_variants_exclude_vcf_df[, c("CHROM", "POS","REF", "ALT")], file = "igv_curated_variants_RM_exclude_vcf_list.tsv", delim = "\t", col_names = TRUE)

```

```{r curated_variants_vcf_or_not}

print(paste0("Manually curated variants to include in new mosaic benchmark vcf: ", nrow(curated_variants_include_vcf_df)))

print(paste0("Manually curated variants to exclude from new mosaic benchmark vcf: ", nrow(curated_variants_exclude_vcf_df)))

```

#### IV. EXCLUDE BED REGIONS

##### Get regions to exclude from new benchmark bed

```{r merge_decision_tree_variants_to_exclude_for_benchmark_bed}

if (!dir.exists("exclude_regions")) {
  dir.create("exclude_regions")
}

# Make df with regions to exclude from benchmark bed (corresponds to decision tree yellow squares)

remove_from_bed_df <- bind_rows(denovo_db_df, easy_NOT_lowerCI5_cand_df, 
                                easy_NOT_lowerCI5_put_pacbio_depth_yes_df, 
                                not_easy_homopol_df, 
                        not_easy_NON_homopol_NOT_pacbio_lowerCI5_pacbio_NOT_upperCI3_df) %>%
                      select(CHROM, POS, REF, ALT) %>%
                      arrange(factor(CHROM, levels = chrom_order), POS)

#Create a new dataframe with the required columns for the BED file (considers both snvs and indels)
exclusions_bed_df <- remove_from_bed_df %>% 
  mutate(start = POS,
         end = POS,
         is_del = nchar(REF) > 1,
         is_ins = nchar(ALT) > 1,
         is_multi_base = nchar(gsub(",", "", ALT)) > 1,
         start = if_else(is_del, POS - 1, start),
         end = if_else(is_del, POS + nchar(REF) - 1, end),
         end = if_else(is_ins, POS + nchar(ALT) - 1, end)) %>%
  select(chrom = CHROM, start, end)

# Write df to BED file
write.table(exclusions_bed_df,
            file = "exclude_regions/decision_tree_regions_to_exclude_RM.bed",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

print(paste0("Decision tree regions to exclude from benchmark bed: ", nrow(exclusions_bed_df)))

```

##### Get GIABv3.1 stratification (AllTandemRepeatsandHomopolymers) and prepare genome file

```{r get_giab_strat_bed}

download.file("https://ftp-trace.ncbi.nlm.nih.gov/ReferenceSamples/giab/release/genome-stratifications/v3.1/GRCh38/LowComplexity/GRCh38_AllTandemRepeatsandHomopolymers_slop5.bed.gz", "exclude_regions/GRCh38_AllTandemRepeatsandHomopolymers_slop5.bed.gz")

system("mamba activate hli_env")

# Index reference
system("samtools faidx GCA_000001405.15_GRCh38_no_alt_analysis_set.fna.fai > GRCh38.fa.fai")

# Prepare genome files with chromosome lengths
system("cat GRCh38.fa.fai | cut -f 1,2 | grep -Ev '^[chr0-9XYM]_|^[chr0-9][0-9XYM]_|^chrUn_' > human.b38.genome")

```

##### Expand bed regions for exclusion from benchmark

```{r expanded_regions_draft_benchmark}

# bedtools commands (intersectBed, multiIntersectBed, mergeBed) to get expanded bed regions

system("intersectBed -wa -a exclude_regions/GRCh38_AllTandemRepeatsandHomopolymers_slop5.bed.gz -b exclude_regions/decision_tree_regions_to_exclude_RM.bed | multiIntersectBed -i stdin exclude_regions/decision_tree_regions_to_exclude_RM.bed | mergeBed -i stdin > exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_RM.bed")

expanded_bed_df <- read.table("exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_RM.bed", sep = "\t")

print(paste0("Expanded bed regions: ", nrow(expanded_bed_df)))
```

##### Slop bed exclusion regions

```{r slop_expanded_bed_regions}

# Add slop (50bp) to each side of regions in expanded bed

system("slopBed -i exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_RM.bed -g human.b38.genome -b 50 > exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_slop50_RM.bed")

expanded_bed_slop_df <- read.table("exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_slop50_RM.bed", sep = "\t")
print(paste0("Expanded bed regions including slop: ", nrow(expanded_bed_slop_df)))
```

##### Create mosaic benchmark bed

```{r exclude_expanded_bed_slop_regions}

# Exclude expanded region of interest w/slop from draft benchmark bed

system("subtractBed -a ~/git_repos/mosaic-benchmarking/data/draft_benchmark/GRCh38_aj-trio_v4.2.1_mosaic-target.bed -b exclude_regions/decision_tree_regions_to_exclude_repeatexpanded_slop50_RM.bed > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.bed")

draft_benchmark_bed <- read.table("~/git_repos/mosaic-benchmarking/data/draft_benchmark/GRCh38_aj-trio_v4.2.1_mosaic-target.bed", sep = "\t")

new_benchmark_bed <- read.table("exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.bed", sep = "\t")

print(paste0("Regions in draft mosaic benchmark bed: ", nrow(draft_benchmark_bed)))
print(paste0("Regions in new v0.0.4 benchmark bed after subtraction: ", nrow(new_benchmark_bed)))
```

##### Get draft mosaic BED summary

```{r bedtools_summary_draft_mosaic_bed}

#<---AMEND to convert to tidy or gt table -->#

# Get draft mosaic bed summary

input_file1 <- "~/git_repos/mosaic-benchmarking/data/draft_benchmark/GRCh38_aj-trio_v4.2.1_mosaic-target.bed"
genome_file <- "human.b38.genome"

draft_bed_summary <- system(paste0("bedtools summary -i ", input_file1, " -g ", genome_file, " > exclude_regions/original_mosaic_benchmark_bed_RM_summary.txt"))

summary_new_bed <- paste("bedtools summary -i", input_file1, "-g", genome_file)  
system(summary_new_bed)

```

##### Get new mosaic BED summary after region exclusion

```{r bedtools_summary_new_mosaic_bed}

# Get new mosaic bed summary

input_file2 <- "exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.bed"

new_bed_summary <- system(paste0("bedtools summary -i ", input_file2, " -g ", genome_file, " > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_bed_RM_summary.txt"))

summary_input_bed <- paste("bedtools summary -i", input_file2, "-g", genome_file)  
system(summary_input_bed)

```

##### Check mosaic benchmark BED for RM benchmark variants

```{r new_mosaic_bed_check_benchmark_vcf}

# Confirm benchmark variants present in bed

#system("bedtools intersect -a vcf_file -b bed_file > out.vcf")

# Check for variants present outside of benchmark regions

```

##### Check mosaic benchmark BED for external validator FP variants

```{r new_mosaic_bed_check_discord_external_FP}

#List of discordant and external FPs generated from external val spreadsheet (18, and 3 of these present in bed but not in vcf - )

# intersectBed -v -a external_validator_discordant_and_FP_snv_list.bed -b ~/Documents/mosaic_variants_project/draft_benchmark/final_vcf_bed_RM/HG002_GRCh38_new_mosaic_benchmark_RM.vcf -v | intersectBed -a stdin -b ~/Documents/mosaic_variants_project/draft_benchmark/final_vcf_bed_RM/HG002_GRCh38_new_mosaic_benchmark_RM.bed | less

```

#####  Exclude short (<50bp) regions and chr6:126382244 and chr10:106867519 from bed

```{r new_mosaic_adjust_bed}

# Review bed summary above
# Confirm no NIST RM benchmark variants occur in regions <50bp 
# Remove short regions  - minimum interval size is 1bp, so need to remove these

system("awk '$3-$2>49' exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.bed > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM_gt49.bed")

# chr6 126382244 - remove from both the vcf and bed
# chr10: 106867519 and surrounding bases so can be treat parents as normal during tumor-normal simulation

system("intersectBed -wa -a exclude_regions/GRCh38_AllTandemRepeatsandHomopolymers_slop5.bed.gz -b exclude_regions/chr6_126382244_chr10_106867519.bed | multiIntersectBed -i stdin exclude_regions/chr6_126382244_chr10_106867519.bed | mergeBed -i stdin > exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_RM.bed")

expanded_bed2_df <- read.table("exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_RM.bed", sep = "\t")

print(paste0("Expanded bed regions: ", nrow(expanded_bed2_df)))

system("slopBed -i exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_RM.bed -g human.b38.genome -b 50 > exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_slop50_RM.bed")

expanded_bed2_slop_df <- read.table("exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_slop50_RM.bed", sep = "\t")
print(paste0("Expanded bed regions including slop: ", nrow(expanded_bed2_slop_df)))

system("subtractBed -a exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM_gt49.bed -b exclude_regions/chr6_126382244_chr10_106867519_regions_to_exclude_repeatexpanded_slop50_RM.bed > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM_gt49_chr6_chr10_variants_removed.bed")

input_file3 <- "exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM_gt49_chr6_chr10_variants_removed.bed"

new_bed_summary <- system(paste0("bedtools summary -i ", input_file3, " -g ", genome_file, " > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM_gt49_chr6_chr10_variants_removed.bed_summary.txt"))

summary_input_bed <- paste("bedtools summary -i", input_file3, "-g", genome_file)  
system(summary_input_bed)

```

##### Get regions to exclude from from draft benchmark VCF

```{r get_regions_to_exclude_from_draft_mosaic_benchmark_vcf}

exclude_from_vcf <- draft_mosaic_db %>%
                    anti_join(curated_variants_include_vcf_df, by = c("CHROM", "POS", "REF", "ALT")) %>%
                    arrange(factor(CHROM, levels = chrom_order), POS)

exclusions_vcf_df <- exclude_from_vcf %>% 
  mutate(start = POS,
         end = POS,
         is_del = nchar(REF) > 1,
         is_ins = nchar(ALT) > 1,
         is_multi_base = nchar(gsub(",", "", ALT)) > 1,
         start = if_else(is_del, POS - 1, start),
         end = if_else(is_del, POS + nchar(REF) - 1, end),
         end = if_else(is_ins & !is_multi_base, POS + nchar(ALT) - 1, end)) %>%
  select(chrom = CHROM, start, end)

write.table(exclusions_vcf_df,
            file = "exclude_regions/exclude_regions_from_vcf_RM.bed",
            sep = "\t",
            quote = FALSE,
            row.names = FALSE,
            col.names = FALSE)

print(paste0("Decision tree variants to exclude from benchmark vcf: ", nrow(exclusions_vcf_df)))

```

##### Create new mosaic benchmark VCF

```{r bedtools_subtract_for_final_benchmark_vcf, warning=FALSE}

# Create new mosaic benchmark vcf via bedtools subtract

system("bedtools subtract -a ~/git_repos/mosaic-benchmarking/data/draft_benchmark/GRCh38_HG002_v0.0.1_mosaic-benchmark.vcf -b exclude_regions/exclude_regions_from_vcf_RM.bed -header > exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.vcf")

```

##### Get draft and new benchmark VCF counts

```{r compare_draft_to_new_benchmark, results = 'hide'}

draft_benchmark_vcf <- read.vcfR("~/git_repos/mosaic-benchmarking/data/draft_benchmark/GRCh38_HG002_v0.0.1_mosaic-benchmark.vcf")

new_mosaic_benchmark_vcf_df <- read.table("exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.vcf", sep = "\t")
```

```{r draft_new_bench_counts}
print(paste0("Draft mosaic benchmark vcf variants: ", nrow(draft_benchmark_vcf)))

print(paste0("Variants in new mosaic benchmark vcf after subtraction: ", nrow(new_mosaic_benchmark_vcf_df)))

```

#### V. DISCORDANCE CHECK

```{r discordance_check, warning=FALSE}

#Generate new mosaic benchmark df
new_vcf <- read.table("exclude_regions/HG002_GRCh38_new_draft_mosaic_benchmark_RM.vcf", header = FALSE, comment.char = "#", sep = "\t")

#Rename columns
new_vcf_df <- select(new_vcf, V1, V2)
colnames(new_vcf_df) <- c("CHROM", "POS")

#Generate df for curated variants to include in vcf
include_in_vcf_df<- select(curated_variants_include_vcf_df, CHROM, POS)

# Get discordant variants from new benchmark vcf and include vcf dataframes  ##this should be 0 
discord_check <- anti_join(new_vcf_df, include_in_vcf_df, by = c("CHROM", "POS"))

# Compare discordant variants to draft mvdb
discord_check_against_mvdb <- inner_join(discord_check, draft_mosaic_db, by = c("CHROM", "POS")) ##this should be 0

print(paste0("Variants in new benchmark vcf that do not match include_in_vcf_df (discordant) : ", nrow(discord_check)))

print(paste0("Discordant variants that hit mvdb: ", nrow(discord_check_against_mvdb)))

```
